Select-based random access to variable-byte encodings

### 2020-03-01 gradu/msc_english_contents.tex
Fixes

An alternative solution with $select$ data structure is proposed and explained in detail later in this thesis. Comparison to $rank$ by Brisaboa et al is shown with different
implementations of $rank$ and $select$ and several different data sets are used. Proposed method does not use data 
reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
new codewords to data when encoding, reading the data straight from the memory is much faster than reassembling the word from blocks.

@@

A good data set for VB encoding is a list of mostly small numbers with a need to support larger ones. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number 
instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. 
In contrast, rare words have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.

@@

Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small numbers, 
which allows them to outperform VB encoding on small number datasets \Citep{Wil99}. Elias Delta and Gamma codes can't support fast random access efficiently because
of their changing code length.

@@

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar encoding types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was was to be found efficient in compressing integer lists. It was first experimented 
as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different 
approaches have been introduced.

@@

\hline
\end{tabular}
\caption{VByte encoded numbers, block size 4. Continuation bit underlined.\label{table:vbytes}}
\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.4]{bris.png}
\caption{Data structure by Brisaboa et al., visualized} (TODO: create this with tikz)\label{bris_ds}
\end{figure}

To decode an element from index i, first block is fetched from $A_1$[i]. Because each is composed of at least one block, the first block is obtainable via straight indexing. With small numbers, this 
is all that is required. If the data has a lot of small numbers, the $rank$ method has a huge advantage. However if the element was stored in multiple blocks, $i \gets rank_0$($A_1$, i) is called. $Rank_0$ 
returns the number of zeros preceding i. In other words, the number of blocks that continue to the next array. This in turn means that $i$ now has the index of the element's next block in array $A_2$.
This process is then repeated until ith bit of the current bit array is set. The decompressed integer is constructed from the fetched blocks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} 
with block length of 8.

\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number \mathbin{|} block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number \mathbin{|} block$
\end{algorithmic}
\end{minipage}
\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}

@@

\begin{itemize}
  \item all - integers equally 1-4 bytes long
  \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
  \item twolarge2 - Same as twolarge, but third of the 1 byte numbers 4 bits or less
  \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
  \item onebyte - all integers 4 bits or less
\end{itemize}

All implementation read the data set from a file and compres it to memory. Then one million index numbers are randomized to an array. The times shown in Table~\ref{table:results1} 
are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.

>>>>

An alternative solution with $select$ data structure is proposed and explained in detail later in this thesis. Comparison to $rank$ by Brisaboa et al is shown with different
implementations of $rank$ and $select$ and several different data sets are used. Proposed method does not use data 
reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
new codewords to data when encoding, data can be read straight from the memory and reassembling the integer from blocks is not needed.

@@

A good data set for VB encoding is a list of mostly small integers with a need to support larger ones. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number 
instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the integers are small. 
In contrast, rare words have only a few entries but the integers stored are larger. These lists are excellent data sets for VB encoding.

@@

Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small integers, 
which allows them to outperform VB encoding on small number datasets \Citep{Wil99}. Elias Delta and Gamma codes can't support fast random access efficiently because
of their changing code length.

@@

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the integers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar encoding types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was was to be found efficient in compressing integer lists. It was first experimented 
as a tool for compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different 
approaches have been introduced.

@@

\hline
\end{tabular}
\caption{VByte encoded integers, block size 4. Continuation bit underlined.\label{table:vbytes}}
\end{table}

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}
[node distance=0pt,
 start chain = A going right,
 arrow/.style = {draw=#1,-{Stealth[]}, 
                shorten >=1mm, shorten <=1mm}, % styles of arrows
 arrow/.default = black,
 X/.style = {rectangle, draw,% styles of nodes in string (chain)
                minimum width=7ex, minimum height=4ex,
                outer sep=0pt, on chain}
 ]
\node[X, join] (start) {$C_{1,2}C_{1,1}$};
\node [left=1ex of start]{\textbf{C =}};
\node[X] {$C_{2,1}$};
\node[X] {$C_{3,3}C_{3,2}C_{3,1}$};
\node[X] {$C_{4,2}C_{4,1}$};
\node[X] {$C_{5,1}$...};

\node[X, below=5ex of start] (a1) {$A_{1,1}$};
\node [left=1ex of a1]{\textbf{$A_1$ =}};
\node[X] {$A_{2,1}$};
\node[X] {$A_{3,1}$};
\node[X] {$A_{4,1}$};
\node[X] {$A_{5,1}$};
\node[X] {...};

\node[X, below=0ex of a1] (b1) {1};
\node [left=1ex of b1]{\textbf{$B_1$ =}};
\node[X] {0};
\node[X] {1};
\node[X] {1};
\node[X] {0};
\node[X] {...};

\node[X, below=5ex of b1] (a2) {$A_{1,2}$};
\node [left=1ex of a2]{\textbf{$A_1$ =}};
\node[X] {$A_{3,2}$};
\node[X] {$A_{4,2}$};
\node[X] {...};

\node[X, below=0ex of a2] (b2) {0};
\node [left=1ex of b2]{\textbf{$B_2$ =}};
\node[X] {1};
\node[X] {0};
\node[X] {...};

\node[X, below=5ex of b2] (a3) {$A_{3,3}$};
\node [left=1ex of a3]{\textbf{$A_3$ =}};
\node[X] {...};

\node[X, below=0ex of a3] (b3) {0};
\node [left=1ex of b3]{\textbf{$B_3$ =}};
\node[X] {...};

\end{tikzpicture}
\caption{Data structure by Brisaboa et al., visualized}\label{bris_ds}

To decode an element from index i, first block is fetched from $A_1$[i]. Because each encoded element is composed of at least one block, the first block is obtainable via straight indexing. With small integers, this 
is all that is required. If the data has a lot of small integers, the $rank$ method has a huge advantage. However if the element was stored in multiple blocks, $i \gets rank_0$($A_1$, i) is called. $Rank_0$ 
returns the number of zeros in the bit array preceding index i. In other words, the number of blocks that continue to the next array. This in turn means that $i$ now has the index of the element's next block in array $A_2$.
This process is then repeated until ith bit of the current bit array is set. The decompressed integer is constructed from the fetched blocks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} 
with block length of 8.

\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number \mathbin{|} block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\end{algorithmic}
\end{minipage}
\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}

@@

\begin{itemize}
  \item all - integers equally 1-4 bytes long
  \item twolarge - one eighth of integers 4 bytes long, one eighth 2 bytes long, rest 1 byte    
  \item twolarge2 - Same as twolarge, but third of the 1 byte integers 4 bits or less
  \item onelarge - one eighth of integers 2 bytes long, rest 4 bits or less
  \item onebyte - all integers 4 bits or less
\end{itemize}

All implementation read the data set from a file and compres it to memory. Then one million index integers are randomized to an array. The times shown in Table~\ref{table:results1} 
are times taken from looping through the index array and VB decoding the number in the index. Additionally, each run is iterated ten times and a mean of the run is taken.
